### 2008-12-16 talk/ecoop2009/jitgen.tex
steal some sections from the EU report

Assume the Python bytecode to be constant, and constant-propagate it into the
Python interpreter.

@@

@@

* Translation time

  - pypy-c-jit is translated into an executable

  - the JIT compiler is automatically generated

* Compile-time: the JIT compiler runs

* Runtime: the JIT compiled code runs

* Compile-time and runtime are intermixed

>>>>

Assume the Python bytecode to be constant, and constant-propagate it into the
Python interpreter.
\cfbolz{note to self: steal bits from the master thesis?}

\cfbolz{I would propose to use either TLC as an example here, or something that
looks at least like an interpreter loop}

@@

Partial evaluation (PE) comes in two flavors:

\begin{itemize}
\item \emph{On-line} partial evaluation: a compiler-like algorithm takes the
source code of the function \texttt{f(x, y)} (or its intermediate representation,
i.e. its control flow graph), and some partial
information, e.g. \texttt{x=5}.  From this, it produces the residual function
\texttt{g(y)} directly, by following in which operations the knowledge \texttt{x=5} can
be used, which loops can be unrolled, etc.

\item \emph{Off-line} partial evalution: in many cases, the goal of partial
evaluation is to improve performance in a specific application.  Assume that we
have a single known function \texttt{f(x, y)} in which we think that the value
of \texttt{x} will change slowly during the execution of our program – much
more slowly than the value of \texttt{y}.  An obvious example is a loop that
calls \texttt{f(x, y)} many times with always the same value \texttt{x}.  We
could then use an on-line partial evaluator to produce a \texttt{g(y)} for each
new value of \texttt{x}.  In practice, the overhead of the partial evaluator
might be too large for it to be executed at run-time.  However, if we know the
function \texttt{f} in advance, and if we know \emph{which} arguments are the
ones that we will want to partially evaluate \texttt{f} with, then we do not
need a full compiler-like analysis of \texttt{f} every time the value of
\texttt{x} changes.  We can precompute once and for all a specialized function
\texttt{f1(x)}, which when called produces the residual function \texttt{g(y)}
corresponding to \texttt{x}.  This is \emph{off-line partial evaluation}; the
specialized function \texttt{f1(x)} is called a \emph{generating extension}.
\end{itemize}

Off-line partial evaluation is usually based on \emph{binding-time analysis}, which
is the process of determining among the variables used in a function (or
a set of functions) which ones are going to be known in advance and
which ones are not.  In the example of \texttt{f(x, y)}, such an analysis
would be able to infer that the constantness of the argument \texttt{x}
implies the constantness of many intermediate values used in the
function.  The \emph{binding time} of a variable determines how early the
value of the variable will be known.

\cfbolz{XXX: unclear how the next paragraph will fit into the text in the end.
it's certainly wrong as it is}
Once binding times have been determined, one possible approach to
producing the generating extension itself is by self-applying on-line
partial evaluators.  This is known as the second Futamura projection
\cite{Futamura99}.  So far it is unclear if this approach can lead to optimal
results, or even if it scales well.  In PyPy we selected a more direct
approach: the generating extension is produced by transformation of the
control flow graphs of the interpreter, guided by the binding times.  We
call this process \emph{timeshifting}.

@@

PyPy contains a framework for generating just-in-time compilers using
off-line partial evaluation.  As such, there are three distinct phases:

\begin{itemize}
\item \emph{Translation time:} during the normal translation of an RPython
program, say the TLC interpreter, we perform binding-time analysis on the
interpreter.  This produces a generating extension, which is linked with the
rest of the program. \cfbolz{XXX not quite right}

\item \emph{Compile time:} during the execution of the program, when a new
bytecode is about to be interpreted, the generating extension is invoked
instead.  As the generating extension is a compiler, all the computations it
performs are called compile-time computations.  Its sole effect is to produce
residual code.

\item \emph{Run time:} the normal execution of the program (which includes the
time spent running the residual code created by the generating extension).
\end{itemize}

Translation time is a purely off-line phase; compile time and run time are
actually highly interleaved during the execution of the program.

\subsection{Binding Time Analysis}

At translation time, PyPy performs binding-time analysis of the source
RPython program after it has been turned to low-level graphs, i.e. at
the level at which operations manipulate pointer-and-structure-like
objects.

The binding-time terminology that we are using in PyPy is based on the
colors that we use when displaying the control flow graphs:

\begin{itemize}
\item \emph{Green} variables contain values that are known at compile-time;
\item \emph{Red} variables contain values that are not known until run-time.
\end{itemize}

The binding-time analyzer of our translation tool-chain is based on the
same type inference engine that is used on the source RPython program,
the annotator.  In this mode, it is called the \emph{hint-annotator}; it
operates over input graphs that are already low-level instead of
RPython-level, and propagates annotations that do not track types but
value dependencies and manually-provided binding time hints.

The normal process of the hint-annotator is to propagate the binding
time (i.e. color) of the variables using the following kind of rules:

\begin{itemize}
\item For a foldable operation (i.e. one without side effect and which depends
only on its argument values), if all arguments are green, then the result can
be green too.

\item Non-foldable operations always produce a red result.

\item At join points, where multiple possible values (depending on control
flow) are meeting into a fresh variable, if any incoming value comes from a red
variable, the result is red.  Otherwise, the color of the result might be
green.  We do not make it eagerly green, because of the control flow
dependency: the residual function is basically a constant-folded copy of the
source function, so it might retain some of the same control flow.  The value
that needs to be stored in the fresh join variable thus depends on which
branches are taken in the residual graph.
\end{itemize}

\subsubsection{Hints}

Our goal in designing our approach to binding-time analysis was to
minimize the number of explicit hints that the user must provide in
the source of the RPython program.  This minimalism was not pushed to
extremes, though, to keep the hint-annotator reasonably simple.

The driving idea was that hints should be need-oriented.  Indeed, in a
program like an interpreter, there are a small number of places where it
would be clearly beneficial for a given value to be known at
compile-time, i.e. green: this is where we require the hints to be
added.

The hint-annotator assumes that all variables are red by default, and
then propagates annotations that record dependency information.
When encountering the user-provided hints, the dependency information
is used to make some variables green.  All
hints are in the form of an operation \texttt{hint(v1, someflag=True)}
which semantically just returns its first argument unmodified.

The crucial need-oriented hint is \texttt{v2 = hint(v1, concrete=True)}
which should be used in places where the programmer considers the
knowledge of the value to be essential.  This hint is interpreted by
the hint-annotator as a request for both \texttt{v1} and \texttt{v2} to be green.  It
has a \emph{global} effect on the binding times: it means that not only
\texttt{v1} but all the values that \texttt{v1} depends on – recursively –
are forced to be green.  The hint-annotator complains if the
dependencies of \texttt{v1} include a value that cannot be green, like
a value read out of a field of a non-immutable structure.

Such a need-oriented backward propagation has advantages over the
commonly used forward propagation, in which a variable is compile-time
if and only if all the variables it depends on are also compile-time.  A
known issue with forward propagation is that it may mark as compile-time
either more variables than expected (which leads to over-specialization
of the residual code), or less variables than expected (preventing
specialization to occur where it would be the most useful).  Our
need-oriented approach reduces the problem of over-specialization, and
it prevents under-specialization: an unsatisfiable \texttt{hint(v1,
concrete=True)} is reported as an error.

In our context, though, such an error can be corrected.  This is done by
promoting a well-chosen variable among the ones that \texttt{v1} depends on.

Promotion is invoked with the use of a hint as well:
\texttt{v2 = hint(v1, promote=True)}.
This hint is a \emph{local} request for \texttt{v2} to be green, without
requiring \texttt{v1} to be green.  Note that this amounts to copying
a red value into a green one, which is not possible in classical
approaches to partial evaluation.  See the Promotion section XXX ref for a
complete discussion of promotion.
